---
title: "Final Project"
author: "Iman Byndloss, Brock Keller, and Rachel Williams"
date: "2025-02-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(dplyr)
library(tidyverse)
library(moments)
library(ggplot2)
library(lubridate)
library(readxl)
library(FSA)
library(lme4)
library(lmerTest)
library(dunn.test)
library(gt)
library(paletteer)
library(tidyr)
library(tibble)
library(ggpubr)
library(viridis)
```


```{r}
#NC air data from EPA
Air_Data_NC <- read.csv("NC_Air_Data_Project.csv")
```

##Data preparation and exploratory analysis

```{r}
#making a plot for first assignment
#annual range in concentrations of select chemical in each site
unique(Air_Data_NC$SITE_ID)
#setting concentrations to be numeric

Air_Data_NC <- Air_Data_NC %>% 
  mutate(across(c(Week, Year, Ca, Cl, HNO3, HNO3.PPB, K, Mg, Na, NH4, NO3, SO2, SO2.PPB, SO4, TNO3,), as.numeric))

#Changing SITE_ID to name of actual town
Air_Data_NC <- Air_Data_NC %>% 
  mutate(SITE_ID = case_when(
    SITE_ID == "BFT142" ~ "Beaufort",
    SITE_ID == "CND125" ~ "Candor",
    SITE_ID == "COW137" ~ "Coweeta",
    SITE_ID == "DUK008" ~ "Duke Forest",
    SITE_ID == "PNF126" ~ "Cranberry"
  ))


#filtering for just 2022
Air_Data_2022 <- Air_Data_NC %>% 
  filter(Year == "2022")

#checking how variables are stored. Looks good
str(Air_Data_2022$SO4)


#grouping by SITE_ID and getting summary statistics
Air_Summary_2022_SO4 <- Air_Data_2022 %>% 
  group_by(SITE_ID) %>% 
  summarize(mean = mean(SO4, na.rm = TRUE),
            sd = sd(SO4, na.rm = TRUE)) %>% 
  ungroup()

#making plot
jpeg(filename = "ByndlossIKellerBWilliamsRFig1.jpeg", 
     width = 2400, height = 1800, res = 300)

SO4_2022_plot <- ggplot()+
  geom_errorbar(data = Air_Summary_2022_SO4, na.rm = TRUE,
                aes(x = SITE_ID, 
                    ymin = mean - sd,
                    ymax = mean + sd,
                    color = SITE_ID),
                width = 0.1, size = 1)+
  geom_jitter(data = Air_Data_2022, na.rm = TRUE,
              aes(x = SITE_ID, y = SO4, 
                  color = SITE_ID),
              alpha = 0.8, size = 0.5)+
  geom_point(data = Air_Summary_2022_SO4, na.rm = TRUE,
             aes(x = SITE_ID, y = mean,
                 color = SITE_ID),
             SIZE = 3)+
  labs(title = "Concentrations of SO4 by Site in 2022",
       x = "Site",
       y = "Weekly Mean SO4 Concentration (ug/m^3)",
       caption = "Figure 1: Figure showing mean weekly SO4 air concentration \n in 2022 across five EPA air monitoring sites in North Carolina.")+
  theme_bw()+ 
  theme(plot.caption = 
      element_text(hjust =  0.5, size = 10))

print(SO4_2022_plot)

dev.off()
```

##Multilevel Regression

```{r}

#Add season collumn, maybe use multilevel model.

#1/Site generates random variable, don't need to specify site

Air_Data_NC$Season <- ifelse(Air_Data_NC$Week %in% c(25:34), "SUMMER",
                    ifelse (Air_Data_NC$Week %in% c(35:51), "AUTUMN",ifelse (Air_Data_NC$Week %in% c(52, 53, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11), 
                                    "WINTER", "SPRING")))

Air_Data_NC$Season <- factor(Air_Data_NC$Season,
                             levels=( c("SUMMER", "AUTUMN", "WINTER", "SPRING")))

Air_Data_NC$SITE_ID <- factor(Air_Data_NC$SITE_ID,
                              levels=(c("Beaufort", "Candor", "Coweeta","Cranberry","Duke Forest")))

#lmer (concentration ~ Season + Year + 1|SITE_ID,
#      data=df)

lmtest <- lm(SO4 ~ Season + Year, data=Air_Data_NC)
summary(lmtest)
plot(lmtest)

ggplot(Air_Data_NC) +
  geom_boxplot(aes(x=SITE_ID, y=SO4)) +
  geom_jitter(aes(x=SITE_ID, y=SO4))


lmm1 <- lmer(SO4 ~ Season + Year + (1|SITE_ID), data=Air_Data_NC)
AIC(lmm1) #Only gas with normal AIC
plot(lmm1)
qqnorm(resid(lmm1),main = "SO4 QQ Plot") 
summary(lmm1) #Lower in Winter and Autumn than in summer. Decreasing over time in the target years. 

lmm2 <- lmer(NO3 ~ Season + Year + (1|SITE_ID), data=Air_Data_NC)
AIC(lmm2)
plot(lmm2)
qqnorm(resid(lmm2),main = "NO3 QQ Plot") 
summary(lmm2)#Autumn, winter, and spring all have significantly higher concentrations than in summer. 

lmm3 <- lmer(HNO3 ~ Season + Year + (1|SITE_ID), data=Air_Data_NC)
AIC(lmm3)
plot(lmm3)
qqnorm(resid(lmm3), main = "HNO3 QQ Plot") 
summary(lmm3) #Shows concentrations are significantly higher in winter and spring than summer

lmm4 <- lmer(NH4 ~ Season + Year + (1|SITE_ID), data=Air_Data_NC)
AIC(lmm4)
plot(lmm4)
qqnorm(resid(lmm4), main = "NH4 QQ Plot") 
summary(lmm4) #Autumn has strong significance, lower concentrations from Summer. It has also been getting lower over the target years. 

par(mfrow = c(2, 2))
qqnorm(resid(lmm1),main = "SO4 QQ Plot") 
qqnorm(resid(lmm4), main = "NH4 QQ Plot") 
qqnorm(resid(lmm2),main = "NO3 QQ Plot")
qqnorm(resid(lmm3), main = "HNO3 QQ Plot")

#Can say which gases have seasonal effects

```

##Linear Regression

```{r}
#Brock's linear regression work

#I want to use week as a proxy for time of year as a predictor of air concentration of whatever gas. I want to convert the week value (1-52) into a sin and cosine wave transformation so that it fits for seasonality and is not treated as a linear variable which would not make sense here (week 52 is close to week 1 and summer is in the middle). 

#Will do like Y = B0 + B1 sin(2pi*week / 52) + B2 cos(2pi*week / 52) + B3 + B4 + E. Thinking to use concentration of another gas as another predictor and also maybe site ID as a factor, though that might be overfitting. 

Air_Data_NC <- Air_Data_NC %>% 
  mutate(
    sin_week = sin(2 * pi * Week / 52),
    cos_week = cos(2 * pi * Week / 52)
  )

ggplot(Air_Data_NC, aes(x = Week)) +
  geom_line(aes(y = sin_week, color = "Sine"), size = 1) +
  geom_line(aes(y = cos_week, color = "Cosine"), size = 1) 


#checking some quick plots to look for correlation in gas concentrations
SO4_NO3 <- ggplot(Air_Data_NC, aes(x = NO3, y = SO4))+
  geom_point()
print(SO4_NO3)
#definitely a decent positive correlation

SO4_HNO3 <- ggplot(Air_Data_NC, aes(x = HNO3, y = SO4))+
  geom_point()
print(SO4_HNO3)
#also pretty decent positive correlation. A little more scattered than previous

SO4_NH4 <- ggplot(Air_Data_NC, aes(x = NH4, y = SO4))+
  geom_point()
print(SO4_NH4)
#really strong correlation!

#trying out a basic first model
lm1 <- lm(SO4 ~ sin_week + cos_week + NH4, data = Air_Data_NC)
summary(lm1)
#this is already a really strong fit and statistically significant

#checking for multi-collinearity
lm_predictors <- Air_Data_NC %>% 
  select(`K`, `Mg`, `Na`,  `NH4`, `NO3`, `SO2`, `SO2.PPB`, `SO4`, `TNO3`, `sin_week`, `cos_week`)

correlation_matrix <- cor(lm_predictors, use = "complete.obs")
correlation_matrix
#this shows that beyond NH4, K and Mg are actually slightly more correlated with SO4 concentration (0.588 and 0.56 respectively). However, it will be important to note that this is correlation alone, and not statistical significance. Some relationship among predictive variables, but not super strong and they will not be included in the regression anyways. 

#trying out a new fit with more variables
lm2 <- lm(SO4 ~ sin_week + cos_week + NH4 + K + Mg, data = Air_Data_NC)
summary(lm2)
#this is way better! all very significant except for sin_week and adj r squared is now 0.77. 
#there are probably other predictors like temperature and population that could be good predictors, but that may be accounted for somewhat already in these pollutant concentrations. 
#lets try removing sin_week just to see. This specific data seems to better align with a cos wave shape, which makes sense given the start in winter and end in winter!

lm3 <- lm(SO4 ~ cos_week + NH4 + K + Mg, data = Air_Data_NC)
summary(lm3)
#this is pretty much the exact same. r squared 0.77 and all significant. I am going to retain sin_week then, as it may make the same model applicable to other datasets and sites whose collection might start and end at other periods of the year. 

#lets check the plots quickly
plot(lm2)
#looks pretty normal, but lets actually run AIC tests instead to determine which model will be the best fit. 

AIC(lm1) #-42
AIC(lm2) #-1122
AIC(lm3) #-1121
#again, the explanatory power of lm2 and lm3 are nearly identical, but I believe lm2 will retain better relevance in other datasets and will be more useful overall. For these reasons, we will select lm2 with predictors of sin_week, cos_week, NH4, K, and Mg as our final linear regression model. 

```


# Spatial Analysis

Before conducting any spatial analysis, it is important to check if the data meets the following parametric assumptions: 1) normally distributed sample values, 2) samples are independent of one another, and 3) population variances are equal.

```{r}
# Filtering data
NC_pollutants <- Air_Data_NC %>%
  select(SITE_ID, HNO3, NH4, NO3, SO4)

# Counting observations per site
bin_info <- NC_pollutants %>%
  group_by(SITE_ID) %>%
  summarize(n = n(),
            bins = ceiling(2 * n^(1/3)))

# Checking for normality through histograms of pollutants by site id
hist1 <- ggplot(NC_pollutants, aes(x = HNO3,
                                   fill = SITE_ID)) +
  scale_fill_viridis(discrete = TRUE, option = "E") +  #colorblindness friendly
  geom_histogram(bins = 13) + #set number of bins according to rice rule
  labs(x = "HNO3 Concentration (µg/m³)",
       y = "Count",
       title = "Distribution of HNO3 Concentration (µg/m³) by Site"#,
       #caption = "Figure 1: This histogram displays the counts for HNO3 
       #concentration, measured in micrograms per cubic 
       #meter, across five sites in North Carolina."
       ) +
  facet_grid(.~SITE_ID) +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))
hist1
hist2 <- ggplot(NC_pollutants, aes(x = NH4,
                                   fill = SITE_ID)) +
  scale_fill_viridis(discrete = TRUE, option = "E") +  #colorblindness friendly
  geom_histogram(bins = 13) + #set number of bins according to rice rule
  labs(x = "NH4 Concentration (µg/m³)",
       y = "Count",
       title = "Distribution of NH4 Concentration (µg/m³) by Site"#,
       #caption = "Figure 2: This histogram displays the counts for NH4 
       #concentration, measured in micrograms per cubic 
       #meter, across five sites in North Carolina."
       ) +
  facet_grid(.~SITE_ID) +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))
hist2
hist3 <- ggplot(NC_pollutants, aes(x = NO3,
                                   fill = SITE_ID)) +
  scale_fill_viridis(discrete = TRUE, option = "E") +  #colorblindness friendly
  geom_histogram(bins = 13) + #set number of bins according to rice rule
  labs(x = "NO3 Concentration (µg/m³)",
       y = "Count",
       title = "Distribution of NO3 Concentration (µg/m³) by Site"#,
       #caption = "Figure 3: This histogram displays the counts for NO3 
       #concentration, measured in micrograms per cubic 
       #meter, across five sites in North Carolina."
       ) +
  facet_grid(.~SITE_ID) +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))
hist3
hist4 <- ggplot(NC_pollutants, aes(x = SO4,
                                   fill = SITE_ID)) +
  scale_fill_viridis(discrete = TRUE, option = "E") +  #colorblindness friendly
  geom_histogram(bins = 13) + #set number of bins according to rice rule
  labs(x = "SO4 Concentration (µg/m³)",
       y = "Count",
       title = "Distribution of SO4 value by Site ID"#,
       #caption = "Figure 4: This histogram displays the counts for SO4 
       #concentration, measured in micrograms per cubic 
       #meter, across five site ids in North Carolina."
       ) +
  facet_grid(.~SITE_ID) +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))
hist4

# Since NO3 non-normally distributed, applying log10 to NO3
NC_pollutants <- NC_pollutants %>%
  mutate(log_NO3 = log10(NO3))

# Creating another histogram with log transformed NO3
hist5 <- ggplot(NC_pollutants, aes(x = log_NO3,
                                   fill = SITE_ID)) +
  scale_fill_viridis(discrete = TRUE, option = "E") +  #colorblindness friendly
  geom_histogram(bins = 13) + #set number of bins according to rice rule
  labs(x = "Log-Transformed NO3 Concentration (µg/m³)",
       y = "Count",
       title = "Distribution of Log-Transformed NO3 Concentration (µg/m³) 
       by Site"#,
       #caption = "Figure 5: This histogram displays the counts for 
       #log-transformed NO3 concentration, measured in 
       #micrograms per cubic meter, across five sites in 
       #North Carolina."
       ) +
  facet_grid(.~SITE_ID) +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))
hist5
```

Based on the histograms, the pollutants are all normally distributed, but it important to note that NO3 had to be log-transformed to meet this assumption.

```{r}
# Conducting Bartlett test for homogeneity
variance_HNO3 <- bartlett.test(NC_pollutants$HNO3,
                               NC_pollutants$SITE_ID)
variance_HNO3
variance_NH4 <- bartlett.test(NC_pollutants$NH4,
                               NC_pollutants$SITE_ID)
variance_NH4
variance_NO3 <- bartlett.test(NC_pollutants$NO3,
                               NC_pollutants$SITE_ID)
variance_NO3
variance_SO4 <- bartlett.test(NC_pollutants$SO4,
                               NC_pollutants$SITE_ID)
variance_SO4
```

Although NO3 could be log-transformed to become more normally distributed, none of the pollutants passed the Barlett test for homogeneity of variances.

```{r}
# Calculating variances of pollutants by site
variance_check_HNO3 <- NC_pollutants %>%
  group_by(SITE_ID) %>%
  summarise(Variance = var(HNO3, na.rm = TRUE))

variance_check_NH4 <- NC_pollutants %>%
  group_by(SITE_ID) %>%
  summarise(Variance = var(NH4, na.rm = TRUE))

variance_check_NO3 <- NC_pollutants %>%
  group_by(SITE_ID) %>%
  summarise(Variance = var(NO3, na.rm = TRUE))

variance_check_SO4 <- NC_pollutants %>%
  group_by(SITE_ID) %>%
  summarise(Variance = var(SO4, na.rm = TRUE))

# Calculating max, min, and max/min for variance 
max_variance_HNO3 <- max(variance_check_HNO3$Variance)
min_variance_HNO3 <- min(variance_check_HNO3$Variance)
variance_ratio_HNO3 <- max_variance_HNO3 / min_variance_HNO3
variance_ratio_HNO3

max_variance_NH4 <- max(variance_check_NH4$Variance)
min_variance_NH4 <- min(variance_check_NH4$Variance)
variance_ratio_NH4 <- max_variance_NH4 / min_variance_NH4
variance_ratio_NH4

max_variance_NO3 <- max(variance_check_NO3$Variance)
min_variance_NO3 <- min(variance_check_NO3$Variance)
variance_ratio_NO3 <- max_variance_NO3 / min_variance_NO3
variance_ratio_NO3

max_variance_SO4 <- max(variance_check_SO4$Variance)
min_variance_SO4 <- min(variance_check_SO4$Variance)
variance_ratio_SO4 <- max_variance_SO4 / min_variance_SO4
variance_ratio_SO4
```

To conduct another test of variance, we calculated the maximum-minimum ratio for variance. If the ratio is less than 4, than equal variances can be assumed. Only two of the four pollutants (SO4 and NH4) passed this test. Due to these results, we will proceed with a one-way ANOVA for SO4 and NH4 but run the non-parametric alternative to a one-way ANOVA, i.e., a Kruskal Wallis test, for HNO3 and NO3.

```{r}
# Running ANOVA + Tukey for NH4 and SO4
anova_NH4 <- aov(NH4 ~ SITE_ID, data = NC_pollutants)
summary(anova_NH4)
tukey_NH4 <- TukeyHSD(anova_NH4)
tukey_NH4

anova_SO4 <- aov(SO4 ~ SITE_ID, data = NC_pollutants)
summary(anova_SO4)
tukey_SO4 <- TukeyHSD(anova_SO4)
tukey_SO4

# Running Kruskal Wallis + Dunn for HO3 and NO3
kw_HO3 <- kruskal.test(HNO3 ~ SITE_ID, data = NC_pollutants)
kw_HO3
dunn_HO3 <- dunn.test(NC_pollutants$HNO3, NC_pollutants$SITE_ID)
dunn_HO3

kw_NO3 <- kruskal.test(NO3 ~ SITE_ID, data = NC_pollutants)
kw_NO3
dunn_NO3 <- dunn.test(NC_pollutants$NO3, NC_pollutants$SITE_ID)
dunn_NO3

# Calculating measures of central tendency and data spread
tapply(NC_pollutants$NH4, NC_pollutants$SITE_ID, mean, na.rm = TRUE)
tapply(NC_pollutants$NH4, NC_pollutants$SITE_ID, median, na.rm = TRUE)
tapply(NC_pollutants$NH4, NC_pollutants$SITE_ID, sd, na.rm = TRUE)
tapply(NC_pollutants$NH4, NC_pollutants$SITE_ID, IQR, na.rm = TRUE)

tapply(NC_pollutants$SO4, NC_pollutants$SITE_ID, mean, na.rm = TRUE)
tapply(NC_pollutants$SO4, NC_pollutants$SITE_ID, median, na.rm = TRUE)
tapply(NC_pollutants$SO4, NC_pollutants$SITE_ID, sd, na.rm = TRUE)
tapply(NC_pollutants$SO4, NC_pollutants$SITE_ID, IQR, na.rm = TRUE)

tapply(NC_pollutants$HNO3, NC_pollutants$SITE_ID, mean, na.rm = TRUE)
tapply(NC_pollutants$HNO3, NC_pollutants$SITE_ID, median, na.rm = TRUE)
tapply(NC_pollutants$HNO3, NC_pollutants$SITE_ID, sd, na.rm = TRUE)
tapply(NC_pollutants$HNO3, NC_pollutants$SITE_ID, IQR, na.rm = TRUE)

tapply(NC_pollutants$NO3, NC_pollutants$SITE_ID, mean, na.rm = TRUE)
tapply(NC_pollutants$NO3, NC_pollutants$SITE_ID, median, na.rm = TRUE)
tapply(NC_pollutants$NO3, NC_pollutants$SITE_ID, sd, na.rm = TRUE)
tapply(NC_pollutants$NO3, NC_pollutants$SITE_ID, IQR, na.rm = TRUE)
```

For NH4, the ANOVA results revealed a significant difference between concentrations across sites in North Carolina (F-value = 33.58, df = 4, p-value < 2e-16). The following post-hoc tukey's HSD test revealed a significant difference between NH4 concentrations across all sites (p < 0.01), except Coweeta-Beaufort (p-value = 1.00), Cranberry-Candor (p-value = 0.56), Duke Forest-Candor (p-value = 0.74), and Duke Forest-Cranberry (p-value = 0.0942623). In terms of the measures of central tendency and spread, for Beaufort, the mean is 0.24, the median is 0.24, the standard deviation is 0.08, and the interquartile range is 0.10. For Candor, the mean is 0.31, the median is 0.29, the standard deviation is 0.12, and the interquartile range is 0.12 For Coweeta, the mean is 0.24, the median is 0.23, the standard deviation is 0.08, and the interquartile range is 0.10. For Cranberry, the mean is 0.29, the median is 0.28, the standard deviation is 0.12, and the interquartile range is 0.12. For Duke Forest, the mean is 0.32, the median is 0.31, the standard deviation is 0.10, and the interquartile range is 0.13.

For SO4, the ANOVA results revealed a significant difference between concentrations across sites in North Carolina (F-value = 74.24, df = 4, p-value < 2e-16). The following post-hoc tukey's HSD test revealed a significant difference between SO4 concentrations across all sites (p < 0.01), except Duke Forest-Candor (p-value = 0.82) and Cranberry-Coweeta (p-value = 0.0923024). In terms of the measures of central tendency and spread, for Beaufort, the mean is 1.06, the median is 1.01, the standard deviation is 0.29, and the interquartile range is 0.38. For Candor, the mean is 0.89, the median is 0.86, the standard deviation is 0.25, and the interquartile range is 0.33. For Coweeta, the mean is 0.67, the median is 0.66, the standard deviation is 0.23, and the interquartile range is 0.33. For Cranberry, the mean is 0.75, the median is 0.68, the standard deviation is 0.29, and the interquartile range is 0.33. For Duke Forest, the mean is 0.91, the median is 0.89, the standard deviation is 0.27, and the interquartile range is 0.34.

For HNO3, the Kruskal-Wallis results revealed a significant difference in concentrations across sites in North Carolina  (chi-squared = 596.6, df = 4, p-value < 2.2e-16). The following Dunn's test reveals that the differences between all sites are significant (p < 0.01). In terms of the measures of central tendency and spread, for Beaufort, the mean is 0.28, the median is 0.26, the standard deviation is 0.13, and the interquartile range is 0.16. For Candor, the mean is 0.54, the median is 0.51, the standard deviation is 0.17, and the interquartile range is 0.24 For Coweeta, the mean is 0.25, the median is 0.23, the standard deviation is 0.11, and the interquartile range is 0.15. For Cranberry, the mean is 0.42, the median is 0.41, the standard deviation is 0.16, and the interquartile range is 0.19. For Duke Forest, the mean is 0.71, the median is 0.67, the standard deviation is 0.25, and the interquartile range is 0.31.

For NO3, the Kruskal-Wallis results revealed a significant difference in concentrations across sites in North Carolina  (chi-squared = 497.75, df = 4, p-value < 2.2e-16). The following Dunn's test reveals that the differences between all sites are significant (p < 0.01). In terms of the measures of central tendency and spread, for Beaufort, the mean is 0.74, the median is 0.70, the standard deviation is 0.25, and the interquartile range is 0.29. For Candor, the mean is 0.37, the median is 0.30, the standard deviation is 0.28, and the interquartile range is 0.31. For Coweeta, the mean is 0.17, the median is 0.14, the standard deviation is 0.12, and the interquartile range is 0.14. For Cranberry, the mean is 0.32, the median is 0.29, the standard deviation is 0.23, and the interquartile range is 0.22. For Duke Forest, the mean is 0.43, the median is 0.37, the standard deviation is 0.28, and the interquartile range is 0.33.
 
```{r}
# Calculating means and standard deviation.
summary_NH4 <- NC_pollutants %>% 
  group_by(SITE_ID) %>%
  summarize(mean = mean(NH4, na.rm = TRUE),
            sd = sd(NH4, na.rm = TRUE)) %>%
  ungroup()

summary_SO4 <- NC_pollutants %>% 
  group_by(SITE_ID) %>%
  summarize(mean = mean(SO4, na.rm = TRUE),
            sd = sd(SO4, na.rm = TRUE)) %>%
  ungroup()
#variance not large and compare median to mean for explanation of why

# Renaming the "mean" column
summary_NH4 <- summary_NH4 %>%
  rename(NH4 = mean)

summary_SO4 <- summary_SO4 %>%
  rename(SO4 = mean)

# Creating plots to display ANOVA results
anova1 <- ggplot() +
  geom_jitter(data = NC_pollutants,
              aes(x = SITE_ID, y = NH4, color = SITE_ID),
              alpha = 0.5, size = 0.5) +
  geom_point(data = summary_NH4,
             aes(x = SITE_ID, y = NH4, color = SITE_ID), 
             size = 3) +
  geom_errorbar(data = summary_NH4,
                aes(x = SITE_ID,
                    ymin = NH4-sd,
                    ymax = NH4+sd,
                    color = SITE_ID),
                width = 0.10, size = 1) +
  scale_color_viridis(discrete = TRUE, option = "E") +  #colorblindness friendly
  labs(x = "Site ID", 
       y = "NH4 Concentration (µg/m³)",
       title = "NH4 Concentration (µg/m³) Across Five Sites in North Carolina: 
       ANOVA Results"#,
       #caption = "Figure 6: This scatter plot with summary statistics
       #displays the NH4 concentration, measured in 
       #micrograms per cubic meter, across five sites 
       #in North Carolina. Individual points represent 
       #observed values while larger points indicate 
       #group means. Error bars show ±1 standard 
       #deviation."
       ) +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))
anova1

anova2 <- ggplot() +
  geom_jitter(data = NC_pollutants,
              aes(x = SITE_ID, y = SO4, color = SITE_ID),
              alpha = 0.5, size = 0.5) +
  geom_point(data = summary_SO4,
             aes(x = SITE_ID, y = SO4, color = SITE_ID), 
             size = 3) +
  geom_errorbar(data = summary_SO4,
                aes(x = SITE_ID,
                    ymin = SO4-sd,
                    ymax = SO4+sd,
                    color = SITE_ID),
                width = 0.10, size = 1) +
  scale_color_viridis(discrete = TRUE, option = "E") +  #colorblindness friendly
  labs(x = "Habitat Type", 
       y = "SO4 Concentration (µg/m³)",
       title = "SO4 Concentration (µg/m³) Across Five Sites in North Carolina: 
       ANOVA Results"#,
       #caption = "Figure 7: This scatter plot with summary statistics
       #displays SO4 concentration, measured in 
       #micrograms per cubic meter, across five 
       #sites in North Carolina. Individual points 
       #represent observed values while larger 
       #points indicate group means. Error bars 
       #show ±1 standard deviation."
       ) +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))
anova2

# Creating plots to display Kruskal Wallis and Dunn results
max_HNO3 <- max(NC_pollutants$HNO3, na.rm = TRUE)
boxplot1 <- ggplot(NC_pollutants, aes(x = SITE_ID, y = HNO3, fill = SITE_ID)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  stat_summary(fun = mean, 
               geom = "point", 
               shape = 23, 
               size = 3, 
               fill = "white") +
  scale_fill_viridis(discrete = TRUE, option = "E") +  #colorblindness friendly 
  theme_minimal() +
  labs(
    title = "Distribution of HNO3 Concentration (µg/m³) by Site",
    x = "Site ID",
    y = "HNO3 Concentration (µg/m³)",
    fill = "Site ID"#,
    #caption = "Figure 8: This boxplot displays the counts for HNO3 
    #concentration, measured in micrograms per cubic 
    #meter, across sites in North Carolina. It includes 
    #annotations for the Kruskal-Wallis and Dunn's tests 
    #results."
    ) +
  theme(legend.position = "none") +
  stat_compare_means(method = "kruskal.test", label = "p.format", 
                     label.x = 2, label.y = max_HNO3+0.4) +
  annotate("text", x = 4, y = max_HNO3+0.4,
           label = "All pairwise comparisons: p < 0.01", 
           size = 4, color = "black", hjust = 0.5) 
boxplot1

max_NO3 <- max(NC_pollutants$NO3, na.rm = TRUE)

boxplot2 <- ggplot(NC_pollutants, aes(x = SITE_ID, y = NO3, fill = SITE_ID)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  stat_summary(fun = mean, 
               geom = "point", 
               shape = 23, 
               size = 3, 
               fill = "white") +
  scale_fill_viridis(discrete = TRUE, option = "E") +  #colorblindness friendly
  theme_minimal() +
  labs(
    title = "Distribution of NO3 Concentration (µg/m³) by Site",
    x = "Site ID",
    y = "NO3 Concentration (µg/m³)",
    fill = "Site ID"#,
    #caption = "Figure 9: This boxplot displays the counts for NO3 concentration,
    #measured in micrograms per cubic meter, across sites in North 
    #Carolina. It includes annotations for the Kruskal-Wallis and 
    #Dunn's tests results."
    ) +
  theme(legend.position = "none") +
  stat_compare_means(method = "kruskal.test", label = "p.format", label.x = 2, label.y = max_NO3+0.4) +
  annotate("text", x = 4, y = max_NO3+0.4,
           label = "All pairwise comparisons: p < 0.01", 
           size = 3, color = "black", hjust = 0.5) 
boxplot2
```

```{r}
# Saving plots
jpeg(filename = "ByndlossIKellerBWilliamsRHist1.jpeg", 
     width = 2400, height = 1800, res = 300)
hist1
dev.off()

jpeg(filename = "ByndlossIKellerBWilliamsRHist2.jpeg", 
     width = 2400, height = 1800, res = 300)
hist2
dev.off()

jpeg(filename = "ByndlossIKellerBWilliamsRHist3.jpeg", 
     width = 2400, height = 1800, res = 300)
hist3
dev.off()

jpeg(filename = "ByndlossIKellerBWilliamsRHist4.jpeg", 
     width = 2400, height = 1800, res = 300)
hist4
dev.off()

jpeg(filename = "ByndlossIKellerBWilliamsRHist5.jpeg", 
     width = 2400, height = 1800, res = 300)
hist5
dev.off()

jpeg(filename = "ByndlossIKellerBWilliamsRAnova1.jpeg", 
     width = 2400, height = 1800, res = 300)
anova1
dev.off()

jpeg(filename = "ByndlossIKellerBWilliamsRAnova2.jpeg", 
     width = 2400, height = 1800, res = 300)
anova2
dev.off()

jpeg(filename = "ByndlossIKellerBWilliamsRBoxplot1.jpeg", 
     width = 2400, height = 1800, res = 300)
boxplot1
dev.off()

jpeg(filename = "ByndlossIKellerBWilliamsRBoxplot2.jpeg", 
     width = 2400, height = 1800, res = 300)
boxplot2
dev.off()
```

# Additional visualizations
```{r}
NC_pollutants_long <- NC_pollutants %>%
  pivot_longer(cols = -c(SITE_ID), 
               names_to = "Pollutant", 
               values_to = "Value")

# Create faceted boxplots
ggplot(NC_pollutants_long, aes(x = SITE_ID, y = Value)) +
  geom_boxplot() +
  theme_minimal() +
  facet_wrap(~ Pollutant, scales = "free_y") +
  labs(title = "Pollutant Concentration (µg/m³) by Site in North Carolina",
       x = "Site ID", y = "Concentration (µg/m³)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```